---
title: "MATH 284/FMPH 291 Survival Analysis  Supplementary Learing Materials"
author: "Yuyao Wang"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tutorial

Here is the link of a tutorial for survival analysis in R by Emily C. Zabor:
[https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html#Part_1:_Introduction_to_Survival_Analysis](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html#Part_1:_Introduction_to_Survival_Analysis)

Part 1 in the above tutorial includes creating survival objects and curve, Kaplan-Meier plots, Estimating x-year survival, estimating median survival time, comparing survival times between groups (using log-rank test), and Cox regression model.

Below are some additional material for different topics.




# Kaplan-Meier Curve

The tutorial uses the '\texttt{ggsurvfit}' R package for doing Kaplan-Meier plots. Alternatively, survival plots can be created using base R or the '\texttt{survminer}' package.
For example, below is an example of fiting KM curves for the two sex groups using the ‘lung’ data set from the '\texttt{survival}’ R package. You may find more details of the usage of the functions (for example how to add confidence intervals) by checking the R document for \texttt{survfit()} and \texttt{plot.survfit()}.

```{r r1_KM}
# Load the survival package
library(survival)
# Fit the Kaplan-Meier model stratified by sex
# The lung dataset is automatically available after loading the survival package
km_fit <- survfit(Surv(time, status) ~ sex, data = lung)
# Plotting the Kaplan-Meier curves for two sex groups
plot(km_fit, main = "Kaplan-Meier Survival Curves by Sex",
        xlab = "Days", ylab = "Survival Probability", col = c(1, 2), lty = 1:2)
# Adding a legend to the plot
legend("topright", legend = c("Male", "Female"), col = c(1, 2), lty = 1:2)
```




# Log-rank test

The function \texttt{survdiff()} used in the above tutorial for comparing survival times between groups is a function that conduct G-rho family tests. The default \texttt{rho=0}, which corresponds to log-rank test. More details can be found in the R documentation of this package.
An alternative way of doing logrank test is using the score test for coxph score test. Try the code below, and you will find the two approaches gives exactly the same results! 

BTW, \texttt{survdiff()} does not handle left truncated data, but `coxph()` can properly handle it by specifying the survival object to be `Surv(Q,X,Delta)`, where Q is the left truncation time, X is the censored event time, and Delta is the event indicator.

```{r r_longrank}
# logrank test using G^\rho test with rho = 0
logrank.1 = survdiff(Surv(time, status) ~ sex, data = lung)
print(logrank.1)

# logrank test using coxph score test
coxphfit = coxph(Surv(time, status) ~ sex, data = lung)
summary(coxphfit)
```







# The dual between estimating T distribution and estimating C distribution

The observed censored event time $X = \min(T,C)$ not only contains information of the event time $T$, but also information of the censoring time $C$. The even indicator $\Delta = I(T<C)$. In other words, when $\Delta = 1$, $X = T$; when $\Delta = 0$, $X = C$. 

When we focus on estimating $T$ distribution, $T$ is censored by $C$, we construct survival object using `Surv(X, delta)`. When we focus on estimating $C$ distribution, $C$ can be viewed as censored by $T$. With this point of view think about how can we construct the survival object.





# Notes on the exponenetial model and weibull model fitted using `survreg()`

The location-scale parameterization of a Weibull distribution found in `survreg` is not the same as the parameterization of rweibull. See the Example for `survreg()` and the chunk for `survreg.distributions` in ‘\texttt{survival}’ R package documentation for more details on how to interpret the outputs.

When fitting an exponential model using `fit = survreg(..., dist = "exponential")`, `1/exp(fit$icoef)` correspond to the rate parameter in `rexp()`. Try the example below.

```{r r_survreg_exp}
n=10000
lambda = 3
TT = rexp(n, rate = lambda)
C = runif(n, 1, 2)
X = pmin(TT,C)
delta = as.numeric(TT<C)

dat = data.frame(X = X, delta = delta)

fit = survreg(Surv(X, delta)~1, data = dat, dist = "exponential")
1/exp(fit$icoef)
```




# Cox proportional hazards regression

We will use the `lung` data set in the `survival` R package as an example.

```{r coxph_data}
summary(lung)
```

## Fitting the model

Try whether you get the same result if you change the data type for `sex` into factor.
How to interpret the results when using `sex` as a continuous variable and when using `sex` as a factor?

```{r coxph}
# lung$sex <- as.factor(lung$sex)  
coxfit = coxph(Surv(time, status) ~ sex, data = lung)
coxfit

ss = summary(coxfit)
ss

# Extract the coefficient 
coef(coxfit)
# or
coxfit$coefficients

# Extract the coefficients and p-value from Wald test
coef(ss)
# or
ss$coefficients
```


## Prediction

Note that when using `basehaz()` to get the estimator of cumulative baseline hazard, one should use the argument `center = FALSE` in order to get the estimate for the cumulative baseline hazard function $\Lambda_0(t)$ in the slides. See the example below.

When using `center = FALSE` in `basehaz()`, one need to further center the covariate values when using the formula:

$$\hat S(t|Z) = e^{-\hat\Lambda_0(t) e^{\hat\beta^\top Z}}.$$

See the example below.

```{r coxph_predict}
coxfit = coxph(Surv(time, status) ~ ph.karno, data = lung)
coxfit

ph.karno.new = 75

## Compute the baseline hazard
baseh = basehaz(coxfit, center = FALSE)
beta = coef(coxfit)

baseh_centerT = basehaz(coxfit, center = TRUE)


newdat = data.frame(time = baseh$time, 
                    status = 1, 
                    ph.karno = ph.karno.new)
                 
pred = predict(coxfit, newdata = newdat, type = "survival", se.fit = TRUE)
pred_surv  = pred$fit
# To see more usage of the function `prediction()` for `coxph` object, try the following.
# ?predict.coxph

# Another method - Compute the estimates from `beta` and `baseh`
# sex = 1 -> to the baseline survival, since sex = 1 is the reference group
pred2_surv_stepf <- stepfun(baseh$time, c(1,exp(-baseh$hazard*exp(beta*ph.karno.new))))
pred3_surv_stepf <- stepfun(baseh_centerT$time, c(1, exp(-baseh_centerT$hazard*exp(beta*ph.karno.new))))
pred4_surv_stepf <- stepfun(baseh_centerT$time, c(1, exp(-baseh_centerT$hazard*exp( beta*(ph.karno.new - mean(lung$ph.karno, na.rm = TRUE)) ))))

## Plot out the step function for the estimated survival curve
pred_surv_stepf <- stepfun(newdat$time, c(1,pred_surv))
plot(pred_surv_stepf, do.points = FALSE, col = 1, lty = 1,
     main = "Predicted survival curves",
     xlab = "Days", ylab = "Survival probability")
# ?plot.stepfun  # for more usage of plotting the `stepfun` object
plot(pred2_surv_stepf, do.points = FALSE, col = 2, lty = 2, add = TRUE)
plot(pred3_surv_stepf, do.points = FALSE, col = 3, lty = 3, add = TRUE)
plot(pred4_surv_stepf, do.points = FALSE, col = 4, lty = 4, add = TRUE)
legend("topright", 
       legend = c("predict.coxph", "center = FALSE, orginal Z", "center = TRUE, orginal Z", "center = TRUE, centered Z"),
       col = 1:4, lty = 1:4,
       bty = "n")
```

See the R documentation for `predict.coxph()` for details about how to do prediction.


Another way to do prediction is using the `hasehaz(.., newdata)`.

```{r coxph_predict2}
newdat = data.frame(ph.karno = ph.karno.new)
cumhazard = basehaz(coxfit, newdata = newdat)

pred5_surv_stepf <- stepfun(baseh$time, c(1, exp(-cumhazard$hazard)))

plot(pred_surv_stepf, do.points = FALSE, col = 1, lty = 1,
     main = "Predicted survival curves",
     xlab = "Days", ylab = "Survival probability")
plot(pred5_surv_stepf, do.points = FALSE, col = 5, lty = 5, add = TRUE)
legend("topright", 
       legend = c("predict.coxph", "basehaz"),
       col = c(1,5), lty = c(1,5),
       bty = "n")


# To predict multiple survival curves at once
ph.karno.new = c(60, 75, 90)
newdat = data.frame(ph.karno = ph.karno.new, sex = c(1,2,1))
cumhazard = basehaz(coxfit, newdata = newdat)
colnames(cumhazard)

surv_stepf <- stepfun(cumhazard$time, c(1, exp(-cumhazard[,1])))
plot(surv_stepf, do.points = FALSE, col = 1, lty = 1,
     main = "Predicted survival curves",
     xlab = "Days", ylab = "Survival probability")
k = nrow(newdat)
for(i in 2:k){
    surv_stepf <- stepfun(cumhazard$time, c(1, exp(-cumhazard[,i])))
    plot(surv_stepf, do.points = FALSE, col = i, lty = i, add = TRUE)
}
legend("topright", 
       legend = paste("ph.karno =",  ph.karno.new),
       col = 1:k, lty = 1:k,
       bty = "n")

```






# Cox PH Model diagnostics

We will use `pbc` data from `survival` R package as illustration.  
Consider the five explanatory variables that are found to be important by Fleming & Harrington (1991):
age, edema, bilirubin, albumin, protime.
As a illustration, we also include an additional binary variable "sex".

```{r}
library(survival)
# library(timereg)

data(pbc, package="survival")
summary(pbc)

covs = c("sex", "age", "edema", "bili", "albumin", "protime")
dat = pbc[, c("id", "time", "status", covs)]

# Remove subjects with missingness
dat = dat[complete.cases(dat), ]
dim(dat)

# View death as event and others as censoring
dat$status <- as.numeric(dat$status == 2)
summary(as.factor(dat$status))

# Change time to years
dat$time = dat$time/365.25


## Summary statistics
summary(dat)
# Histogram for the covariates
# covs.hist = c("age", "albumin", "protime")
covs.hist = c("age", "bili", "albumin", "protime")
par(mfrow=c(2,2))
for(i in 1:length(covs.hist)){
    hist(dat[,covs.hist[i]], xlab = covs.hist[i], main = "")
}



dat$logAlbumin = log(dat$albumin)
dat$logProtime = log(dat$protime)
dat$logBilirubin = log(dat$bili)

summary(dat)

covs.hist = c("age", "logBilirubin", "logAlbumin", "logProtime")
par(mfrow=c(2,2))
for(i in 1:length(covs.hist)){
    hist(dat[,covs.hist[i]], xlab = covs.hist[i], main = "")
}
```


### Fit univariate Cox models

```{r}
# Fit a univariate model for sex
fit <-  coxph(Surv(time,status)~sex,
              data = dat)
fit
summary(fit)


# Fit a univariate model for age
fit<- coxph(Surv(time,status)~age,
            data = dat)
fit
summary(fit)
```


### Fit a multivariate Cox-PH model 

```{r}
fit_multi <- coxph(Surv(time,status)~sex+age+edema+logBilirubin+logAlbumin+logProtime,
             data = dat)
fit_multi
summary(fit_multi)
```


### Compute the R-square meansure based on the partial likelihood ratio statistic under the Cox model

```{r}
library(CoxR2)
coxr2fit = coxr2(fit_multi)
coxr2fit$rsq
```


## Check the PH assumption using cumulative martingale residuals

Here we refit the Cox model with `timereg:cox.aalen()`, which has cox regression as a special case. This package gives us easy implementation for diagnostic plots and tests of proportional hazards assumption using cumulative martingale residuals.

```{r}
library(timereg)
fit_multi2 <- cox.aalen(Surv(time,status)~prop(sex)+prop(age)+prop(edema)+
                            prop(logBilirubin)+prop(logAlbumin)+prop(logProtime),
                        data = dat)
fit_multi2

# p-values from test of PH assumption
summary(fit_multi2)

par(mfrow=c(2,3))
plot(fit_multi2, score = TRUE)
```


## Check the PH assumption using Schoenfeld residuals

```{r}
fit_cox = coxph(Surv(time,status)~sex+age+edema+logBilirubin+logAlbumin+logProtime,
                data = dat)
test.ph = cox.zph(fit_cox)
print(test.ph)
plot(test.ph)
```

We can also do a graphical diagnostic using the function `ggcoxzph()` (in the `survminer` R package), which produces, for each covariate, graphs of the scaled Schoenfeld residuals against the time.
```{r}
library(survminer)
ggcoxzph(test.ph)
```


### Remarks

 - Comparing the test using cumulative martingale residuals and the one using Schoenfeld residuals, we see that the test using Schoenfeld residuals seems to be less sensitive to violations of proportional hazards assumption compared with the test using cumulative martingale residuals.




# Stratified Cox model

Here we fit a stratified Cox-PH model with strata defined by edema. Note that edema only takes 3 values: 0, 0.5, 1. 

```{r}
summary(as.factor(dat$edema))

fit_strata <- coxph(Surv(time,status)~sex+age+strata(edema)+logBilirubin+logAlbumin+logProtime,
                    data = dat)
fit_strata   

## The cumulative baseline hazards for the stratified Cox model
bhaz <- basehaz(fit_strata, centered = FALSE)  # Get baseline hazards

# Plot
par(mfrow=c(1,1))
plot(bhaz$time[bhaz$strata=="edema=0"], bhaz$hazard[bhaz$strata=="edema=0"], 
     type = "s", col = 1, lty = 1, 
     xlab = "Time", ylab = "Cumulative Hazard", 
     ylim = range(bhaz$hazard),
     main = "Cumulative Baseline Hazard by Edema Strata")
lines(bhaz$time[bhaz$strata=="edema=0.5"], bhaz$hazard[bhaz$strata=="edema=0.5"], 
      type = "s", col = 2, lty = 2)
lines(bhaz$time[bhaz$strata=="edema=1"], bhaz$hazard[bhaz$strata=="edema=1"],
      type = "s", col = 3, lty = 3)
legend("topleft", legend = c("Edema = 0", "Edema = 0.5", "Edema = 1"),
       col = 1:3, lty = 1:3, bty = "n", title = "Strata")
```

Here, `type = "s"` in `plot()` specifies that the line should be drawn as a step function. This is an easier way of doing plots for step function than the way before (first define a step function and then do the plot).



# Cox model with piecewise constant time-varying coefficient

We fit a 3 piece piecewise constant model where the time intervals are chosen such that eahc interval has approximately the same number of events.

We use the approach mentioned in the lecture of fitting the equivalent Cox model with time varying covariates. 

The following is one way of creating the pseudo data and fitting a piecewise model utilizing `survival::survSplit()` function.

```{r}
qtl = quantile(dat$time[dat$status==1], c(1/3, 2/3))
qtl = as.numeric(qtl)
qtl

dat_cut = survSplit(Surv(time,status)~., cut = qtl,
                     data = dat, episode = "tgroup")   
# This gives the same pseudo data as in the approach we saw last time.

# Fit the piecewise Cox model
fit_3p = coxph(Surv(tstart,time,status) ~ strata(tgroup):(sex+age+edema+logBilirubin+logAlbumin+logProtime),
               data = dat_cut, id = id)
fit_3p
```


Below we create the pseudo data ourselves without using the `survival::survSplit()` function. We can see that the results are the same as before, which is expected.


```{r}
# pick cuts
qtl = quantile(dat$time[dat$status==1], c(1/3, 2/3))
qtl
qtl = as.numeric(qtl)
qtl

covs = c("sex", "age", "edema", "logBilirubin", "logAlbumin", "logProtime")
dat$sex = as.numeric(dat$sex== "f")   # 1: female, 0: male

# Create pseudo data with time-varying covariates Z*I_{(cut1, cut2]}(t)
dat2 = dat[dat$time> qtl[1], ]
dat3 = dat[dat$time> qtl[2], ]
covs0_1 = matrix(0, nrow = nrow(dat), ncol = length(covs))
covs0_2 = matrix(0, nrow = nrow(dat2), ncol = length(covs))
covs0_3 = matrix(0, nrow = nrow(dat3), ncol = length(covs))

covsname1 = paste(covs, "1", sep = "")
covsname2 = paste(covs, "2", sep = "")
covsname3 = paste(covs, "3", sep = "")

dat_pseudo1 = cbind(id = dat[,"id"], start = 0, stop = pmin(dat$time, qtl[1]), 
                   delta = as.numeric(dat$time<=qtl[1])*dat$status,
                   dat[, covs], covs0_1, covs0_1)
colnames(dat_pseudo1)[5:22] <- c(covsname1, covsname2, covsname3)

dat_pseudo2 = cbind(id = dat2[,"id"], start = qtl[1], stop = pmin(dat2$time, qtl[2]), 
                    delta = as.numeric(dat2$time<=qtl[2])*dat2$status,
                    covs0_2, dat2[, covs], covs0_2)
colnames(dat_pseudo2)[5:22] <- c(covsname1, covsname2, covsname3)

dat_pseudo3 = cbind(id = dat3[,"id"], start = qtl[2], stop = dat3$time, 
                    delta = dat3$status,
                    covs0_3, covs0_3, dat3[, covs])
colnames(dat_pseudo3)[5:22] <- c(covsname1, covsname2, covsname3)

dat_pseudo = rbind(dat_pseudo1, dat_pseudo2, dat_pseudo3)


# Change the order of roles - put the rows for same subjects together
order = order(as.numeric(dat_pseudo$id))
dat_psd = dat_pseudo[order, ]

# colnames(dat_psd)[5:22] <- c(covsname1, covsname2, covsname3)

dat_psd[1:10,]
head(dat)


# Fit the Cox model
fit_3piece = coxph(as.formula(paste("Surv(start, stop, delta)~",
                                    paste(c(covsname1, covsname2, covsname3), collapse = "+"))),
                   data = dat_psd, id = id)

fit_3piece
```


### Remarks

 - The Schoenfeld resiudals plot for `logProtime` shows a downward trend, indicating that $\beta(t)$ is decreasing over time. This coincides with the results from the 3-piece piecewise constant Cox model, which shows the coefficient for `logProtime` on the three time intervals are decreasing.



